
Coursera: Reproducible Research - Course Project 1
==================================================

# Loading the data and getting an overview

```{r readindata}
activityData <- read.csv("activity.csv", header = TRUE)
head(activityData)
summary(activityData)
str(activityData)
```

# Transforming date variable from character to date

```{r}
activityData$date <- as.Date(activityData$date)
```

## Calculate total number of steps taken per day

```{r, echo = FALSE}
library(reshape2) # to get melt and dcast functions
library(dplyr)
```

We melt the data frame to cast by date - the id variable is set to date and the measure variable is set to steps. This way, we get a table with values for steps taken per day

```{r total # of steps taken per day}
activityMeltDate <- melt(activityData, id.vars="date", measure.vars="steps", na.rm=FALSE)
```

Next, we cast the data frame to see the steps per day (in a table with three columns)

```{r}
activityCastDate <- dcast(activityMeltDate, date ~ variable, sum)
```

# Plot histogram of total steps per day

```{r histogram}
library(ggplot2)
stepshist <- ggplot(activityCastDate, aes(date, steps))
stepshist + geom_bar(stat = "identity", fill = "blue") +
  ggtitle("Total steps taken per day/date") +
  xlab("Total steps") +
  ylab("Frequency") +
  geom_hline(aes(yintercept=mean(steps)), col = "red")
```

# Calculate mean and median number of steps taken each day

```{r}
mean_steps <- round(mean(activityCastDate$steps, na.rm = TRUE),2)
mean_steps
```

```{r}
median_steps <- round(median(activityCastDate$steps, na.rm = TRUE),2)
median_steps
```

# Create a time series plot of the average number of steps taken

```{r}
interval<- activityData%>%
        group_by(interval)%>%
        filter(!is.na(steps))%>%
        summarise(avg_steps = mean(steps, na.rm=TRUE))
interval
```

Now, we plot the time series

```{r}
time_series <- ggplot(interval, aes(interval, avg_steps))
time_series +
  geom_line(col = "blue", lwd = 1) + 
  ggtitle("Time series of average steps taken") +
  xlab("Interval") +
  ylab("Average steps") +
  geom_hline(aes(yintercept=mean(avg_steps)), col = "red")
```

# Get the 5-minute interval that contains the maximum number of steps
```{r}
max_steps <- interval[which.max(interval$avg_steps),]
max_steps
```

## Getting and imputing missing data - we are just going to use the mean values

How many missing values are there?

```{r}
missing_values <- sum(is.na(activityData$steps))
paste(missing_values)
```

Since there are 2,304 missing values in the steps variable, we are going to use the mean value for each day to replace the NAs.

```{r}
activityData2<- activityData
missing_values2 <- is.na(activityData2$steps)
avg_interval<- tapply(activityData2$steps, activityData2$interval, mean, na.rm=TRUE)
activityData2$steps[missing_values2] <- avg_interval[as.character(activityData2$interval[missing_values2])]
names(activityData2)
head(activityData2)
```

Are there any missing values left?

```{r}
na_left <- sum(is.na(activityData2))
paste(na_left)
```
No!

# Create a new histogram with the imputed values

This time with a dplyr approach for a change

```{r}
totalSteps<- activityData2%>%
        group_by(date)%>%
        summarise(total_steps = sum(steps, na.rm=TRUE))
```

# Plot new histogram of total steps per day

```{r histogram2}
library(ggplot2)
stepshist2 <- ggplot(totalSteps, aes(date, total_steps))
stepshist2 + geom_bar(stat = "identity", fill = "blue") +
  ggtitle("Total steps taken per day/date without NAs") +
  xlab("Total steps") +
  ylab("Frequency") +
  geom_hline(aes(yintercept=mean(total_steps)), col = "red")
```

# Do the mean and median values differ?

# Calculate mean and median number of steps taken each day

```{r}
mean_steps2 <- round(mean(totalSteps$total_steps, na.rm = TRUE),2)
mean_steps2
```

```{r}
median_steps2 <- round(median(totalSteps$total_steps, na.rm = TRUE),2)
median_steps2
```

With the imputed data included, both the mean and the median are the same now, 10766.19!

## Create a panel plot comparing steps taken per 5-minute interval across weekdays and weekends

```{r}
activityData2$daytype <- ifelse(weekdays(activityData2$date) == "Saturday"|weekdays(activityData2$date) == "Sunday", "weekend", "weekday")
head(activityData2)
```

Time series comparing the average number of steps/5-minute interval across weekdays and weekends
```{r}
interval2<- activityData2%>%
        group_by(interval, daytype)%>%
        filter(!is.na(steps))%>%
        summarise(avg_steps2 = mean(steps, na.rm=TRUE))
interval2
```

```{r}
time_series2 <- ggplot(interval2, aes(interval, avg_steps2, col = daytype))
time_series2 +
  geom_line() + 
  facet_grid(daytype~.) +
  scale_color_manual(values = c("tomato", "steelblue"), name = "Daytype") +
  ggtitle("Time series of average steps taken weekday/weekend") +
  xlab("Interval") +
  ylab("Average steps")
```

